---
title: "Max Temp"
author: "Mikkel Bjornson"
date: "10/17/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

## Library
library(readr)
library(tidyverse)
library(gridExtra)
library(TSA)
library(forecast)

## Data
train <- readRDS(here::here("train.rds"))
test <- readRDS(here::here("test.rds"))
```


```{r}
tmax.p1<-ggplot(train, aes(time, t_max ))+
  geom_line(na.rm = T)+
  theme_classic()+
  geom_smooth(method = "lm")+
  geom_hline(yintercept = mean(train$t_max), color="red", alpha=0.8)

tmax.p2<-ggplot(train, aes(mo, t_max, group=yr, color = -yr))+
  geom_line(na.rm = T, alpha=0.2)+
  scale_x_continuous(breaks = 1:12, labels = month.abb)+
  theme_classic()

tmax.p1
tmax.p2
```

Examining the plots, there appears to be a strong seasonality and a potential upward trend in the data. Visually there does not appear to be a change in the variaiton. As expected, there appears to be an increase in temps in the summer with a peak around August. 

```{r}
## create time series
tmax.train<- ts(train$t_max, start = c(train$yr[1],train$mo[1]), frequency = 12)
tmax.test<- ts(test$t_max, start = c(test$yr[1],test$mo[1]), frequency = 12)

## estimate trend
tmax.trend<- lm(t_max~time, data=train)
tmax.trend.ts<- ts(tmax.trend$fitted.values, start = c(train$yr[1],train$mo[1]), frequency = 12)

## detrend
tmax.train.det <- tmax.train-tmax.trend.ts
plot(tmax.train.det)

## estimate seasonality
mo<- as.factor(train$mo)
tmax.sea<- lm(tmax.train.det~mo)
tmax.sea.ts<- ts(tmax.sea$fitted.values, start = c(train$yr[1],train$mo[1]), frequency = 12)

## de-season
tmax.train.rand<- tmax.train.det-tmax.sea.ts
plot(tmax.train.rand)
```


```{r}
### acf and PACF
acf(tmax.train.rand)
pacf(tmax.train.rand)
eacf(tmax.train.rand)
```

ACF suggests q= 2,3, PACF suggests p = 1. EACF suggests p=0,1 and q = 1,2. Will attempt MA(2), ARMA(1,1), ARMA(1,2), ARMA(1,3).

```{r}
tmax.m1<-arima(tmax.train.rand, order = c(0,0,2), include.mean = F)
tmax.m2<-arima(tmax.train.rand, order = c(1,0,1), include.mean = F)
tmax.m3<-arima(tmax.train.rand, order = c(1,0,2), include.mean = F)
tmax.m4<-arima(tmax.train.rand, order = c(1,0,3), include.mean = F)
tmax.m1
tmax.m2
tmax.m3
tmax.m4
```

The ARMA(1,1) model appears to fit the best. 

```{r}
tsdiag(tmax.m2)
pacf(tmax.m2$residuals)
acf(tmax.m2$residuals)

qqnorm(tmax.m2$residuals)
qqline(tmax.m2$residuals)
```

With the exception of one large outlier the residuals look good. The ACF, PACF, and Ljung-box stats give no reason for concern. The residuals appear to closely follow a normal distribution. 


```{r}
## make predictions
tmax.m2.preds <- predict(tmax.m2, n.ahead = 180)

## estimate seasonality
tmax.sea.pred<- predict(tmax.sea, newdata = data.frame(mo=as.factor(test$mo)))

## estimate trend
tmax.trend.pred<- predict(tmax.trend, newdata = test)

## total
tmax.arma.pred<- tmax.m2.preds$pred + tmax.sea.pred + tmax.trend.pred

ggplot(train, aes(time, t_max))+
  geom_line(na.rm = T)+
  geom_line(data = data.frame(t = test$time, preds=tmax.arma.pred), aes(x=t, y=preds), col='red')+
  labs(title="ARMA(1,1)")
```

```{r}
tmax.arma.rmse<-(mean((tmax.arma.pred-test$t_max)^2))^0.5
```

This method produced forecasts that are 3 degrees off on average. We will attempt to use a SARIMA model to improve those predictions. 



```{r}
tmax.diff<- diff(tmax.train, lag = 1)

ggplot(data.frame(t = tmax.diff, yr = time(tmax.diff)), aes(yr,t))+
         geom_line()+
  geom_smooth(method = "lm")+
  geom_hline(yintercept = mean(tmax.diff), color="red", alpha=0.5)

tmax.diff.sea <- diff(tmax.diff, lag = 12)

ggplot(data.frame(t = tmax.diff.sea, yr = time(tmax.diff.sea)), aes(yr,t))+
         geom_line()

```

A single difference appears to remove the trend suggesting d=1. A single difference also appears to remove the seasonality suggesting D=1. 

```{r}
acf(tmax.diff.sea)
pacf(tmax.diff.sea, lag.max = 48)
eacf(tmax.diff.sea)
```
Season model appears to be MA(1) and random component also appears to be MA(1). Will attempt:
Sarima(0,1,1)x(0,1,1)12
Sarima(1,1,1)x(0,1,1)12
Sarima(1,1,2)x(0,1,1)12

```{r}
tmax.fit1<- arima(tmax.train, order=c(0,1,1), seasonal=list(order=c(0,1,1), period=12))
tmax.fit2<- arima(tmax.train, order=c(1,1,1), seasonal=list(order=c(0,1,1), period=12))
tmax.fit3<- arima(tmax.train, order=c(1,1,2), seasonal=list(order=c(0,1,1), period=12))
tmax.fit4<- arima(tmax.train, order=c(0,1,2), seasonal=list(order=c(0,1,1), period=12))
tmax.fit1
tmax.fit2
tmax.fit3
tmax.fit4
```

Sarima(1,1,2)x(0,1,1)12 appears to fit the model the best.

```{r}
tsdiag(tmax.fit3)
acf(tmax.fit3$residuals)
pacf(tmax.fit3$residuals)

qqnorm(tmax.fit3$residuals)
```

```{r}
tmax.sarima.preds<- predict(tmax.fit3, n.ahead = 180)

tmax.sarima.preds.df<- data.frame(pred = tmax.sarima.preds$pred, 
           lwr= tmax.sarima.preds$pred - 1.965*tmax.sarima.preds$se,
           upr= tmax.sarima.preds$pred + 1.965*tmax.sarima.preds$se,
           time = time(tmax.test))

ggplot(tmax.sarima.preds.df, aes(time, pred))+
  geom_line(data=test, aes(time, t_max))+
  geom_ribbon(aes(x=time, ymin=lwr, ymax=upr), fill="darkgrey", alpha=0.5)+
  geom_line(color="blue")+
  geom_line(data=train, aes(time, t_max))+
  scale_x_continuous(limits = c(2000,2020))
  
```

```{r}
tmax.sarima.rmse<- mean((tmax.sarima.preds$pred - test$t_max)^2)^0.5

tmax.arma.rmse
tmax.sarima.rmse
```

```{r}
nahead<- 12*15
tmax.hw<- hw(tmax.train, seasonal= "additive", h=nahead)

tmax.hw.est<- cbind(test$time, tmax.hw$mean, tmax.hw$upper, tmax.hw$lower) %>% as.data.frame()
names(tmax.hw.est)<-c("time1", "est", "u80", "u95", "l80", "l95")

ggplot(tmax.hw.est, aes(x= time1 , y=est))+
  geom_line()+
  geom_line(data=train, aes(x=time, y= t_max))

tmax.hw$model
```

```{r}
tmax.hw.rmse<- (mean((tmax.hw$mean - test$t_max)^2))^0.5

tmax.hw.rmse
tmax.arma.rmse
tmax.sarima.rmse
```











































